---
title: "Create Clean Data"
author: "Your Name"
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook creates a clean crime table covering the years 1946-2017 for the City of St. Louis. It uses a combination of data from the FBI to produce these estimates. Data prior to 1985 was obtained by entering values from the `.pdf` files included in `data/raw/pdf/` - these are the historic, yearly reports produced by the FBI. From 1985-2014, the data were compiled for an earlier project (will be updating this workflow to use original data sources at a later date). From 2015 onward, there are individual raw data files downloaded from the FBI for each city in Missouri. 

These are combined with historic population data to produce yearly trends for counts and per capita rates for what are known as "index crimes." These are sometimes called Part 1 crimes and are subdivided as follows:

> index crimes... are broken into two categories: violent and property crimes. Aggravated assault, forcible rape, murder, and robbery are classified as violent while arson, burglary, larceny-theft, and motor vehicle theft are classified as property crimes. ([Wikipedia](https://en.wikipedia.org/wiki/Uniform_Crime_Reports))

Beginning in 1980, the FBI began releasing their data with population counts. Prior to this point, population was not included. Data from the [Missouri Census Data Center](http://mcdc.missouri.edu), which is available for decennial census years from 1900 to 2000, was used for years prior to 1980. A model that assumes linear population change was used to estimate population for each year - a decade where 100,000 people were lost in total population would therefore be modeled as having lost 10,000 people each year. All rates included are per 100,000 residents.

In addition to missing the original data for 1985-2014, arson data is missing for those years. Arson was not added as an index crime until 1979. To keep values standard, property crime rates are not calculated for years prior to 1979. Instead, `theftRate` and `theftTotal` values are produced for all years (1949 and later - 1946-1948 had incomplete larceny data) that encompasses burglary, larceny, and theft of a motor vehicle. For years when arson data is available, `propertyRate` and `propertyTotal` values are also calculated.

For violent crime, rape data was not available until 1958. `violentRate` and `violentTotal` values are therefore available for the years 1958 onward. A `murderRate` value is available for all years, and additional rates can be calculated by dividing the count of crimes per year by the total population, and then multiplying that by 100,000. 

## Dependencies
### Packages
This notebook requires a number of `tidyverse` packages plus `here` and `janitor` for data import and wrangling:

```{r load-packages}
# tidyverse packages
library(dplyr)        # data wrangling
library(readr)        # data import
library(readxl)       # data import
library(stringr)      # string tools
library(tidyr)        # data wrangling

# other packages
library(here)         # file paths
library(janitor)      # data wrangling
```

### Functions
To standardize the data cleaning process, a total of six functions have been written. They do not have any additional dependencies beyond those listed above:

```{r}
# calculate totals for crime categories
source(here("source", "violentTotal.R"))
source(here("source", "theftTotal.R"))
source(here("source", "propertyTotal.R"))

# cleaning contemporary (2015 onward) FBI releases
source(here("source", "contemp.R"))

# calculation population change 1946-1980
source(here("source", "popCalc.R"))
```

## Load Data
This notebook requires a number of different data sets:

```{r load-data, echo=FALSE, message=FALSE}
# historic crime data
stLouis46 <- read_csv(here("data", "raw", "tables", "rawFBIData-STL-1946.csv"))
stLouis85 <- read_csv(here("data", "raw", "tables", "rawFBIData-STL-1985.csv"))

# contemporary crime data
stLouis15 <- read_excel(here("data", "raw", "tables", "rawFBIData-STL-2015.xls"), 
                        sheet = 1, col_names = TRUE, col_types = NULL, na = "", 
                        skip = 4)
stLouis16 <- read_excel(here("data", "raw", "tables", "rawFBIData-STL-2016.xls"), 
                        sheet = 1, col_names = TRUE, col_types = NULL, na = "", 
                        skip = 4)
stLouis17 <- read_excel(here("data", "raw", "tables", "rawFBIData-STL-2017.xls"), 
                        sheet = 1, col_names = TRUE, col_types = NULL, na = "", 
                        skip = 4)

# historic population data
stLouisPop <- read_excel(here("data", "raw", "tables", "moco_totpop_1900_2000.xls"),
                         sheet = 1, col_names = TRUE)
```

## Standardize Names and Data
### 1946-1984
The `stLouis46` data were digitized manually using the preferred variable names, thus the only work that remains for these years is to calculate totals for violent crimes, thefts, and property crimes:

```{r}
stLouis46 %>%
  violentTotal() %>%
  theftTotal() %>%
  propertyTotal() -> stLouis46
```

### 1985-2014
For the `stLouis85` data, we need to rename columns and reduce the rape counts from two columns to one:

```{r}
stLouis85 %>% 
  clean_names() %>%
  select(-months, -violent_crime_total, -property_crime_total) %>%
  rename(
    murder = murder_and_nonnegligent_manslaughter,
    rape = legacy_rape_1,
    rape2 = revised_rape_2,
    agAssault = aggravated_assault,
    larceny = larceny_theft,
    mvLarceny = motor_vehicle_theft
  ) %>%
  mutate(rape = ifelse(is.na(rape) == TRUE, rape2, rape)) %>%
  select(-rape2) %>%
  mutate(arson = as.numeric(arson)) -> stLouis85
```

Once we have these data tidied, we can calculate new violent and property crime totals:

```{r}
stLouis85 %>%
  violentTotal() %>%
  theftTotal() %>%
  propertyTotal() -> stLouis85 
```

The arson data is currently missing for this time period.

### 2015-2017
The data for these years are cleaned in two stages - we clean up names so that each table can be combined together with the others, and then calculate crime totals on the single, combined table.

#### 2015 Names
For `stLouis15`, we first need to subset the data so that we only have data on St. Louis and clean up column names:

```{r}
stLouis15 %>% 
  contempNames(year = 2015) %>%
  mutate(year = 2015) %>%
  contempSubset(year = 2015) -> stLouis15
```

#### 2016 Names
The same needs to be done for 2016:

```{r}
stLouis16 %>% 
  contempNames(year = 2016) %>%
  mutate(year = 2016) %>%
  contempSubset(year = 2016) -> stLouis16
```

#### 2017 Names
And for 2017:

```{r}
stLouis17 %>% 
  contempNames(year = 2017) %>%
  mutate(year = 2017) %>%
  contempSubset(year = 2017) -> stLouis17
```

### Combine
Once we have these datas' names standardized, we can bind them together and remove the intermediate objects:

```{r}
# bind tables
stLouis15 <- bind_rows(stLouis15, stLouis16, stLouis17)

# remove intermediate objects
rm(stLouis16, stLouis17)
```

### Crime Counts
With the data combined, we can calculate totals for violent crime, theft, and property crimes:

```{r}
stLouis15 %>%
  violentTotal() %>%
  theftTotal() %>%
  propertyTotal() -> stLouis15
```

## Combine
With all columns standardized, we can bind all of the data together and again remove intermediate objects:

```{r}
# bind tables
stLouis <- bind_rows(stLouis46, stLouis85, stLouis15)

# remove intermediate objects
rm(stLouis46, stLouis85, stLouis15)
```

## Create Population Estimates
Since years prior to 1980 lack population estimates, we need to produce yearly population estimates. To start, we'll clean-up the Missouri historical census data so that we only have data for St. Louis for decennial census years 1930 through 1980. We also reshape the data to long to facilitate additional cleaning:

```{r}
stLouisPop %>%
  filter(fipco == "29510") %>%
  select(-c(fipco, areaname)) %>%
  gather(key = "year", value = "population") %>%
  mutate(year = as.numeric(year)) %>%
  filter(year > 1930 & year < 1990) -> stLouisPop
```

With the data clean, we calculate linear population change estimates for each year between decennial census years. We then combine them all together and insert them into the crime table before removing the intermediate population objects:

```{r}
# calculate yearly population estimates
pop40 <- popCalc(stLouisPop, start = 1940, end = 1950)
pop50 <- popCalc(stLouisPop, start = 1950, end = 1960)
pop60 <- popCalc(stLouisPop, start = 1960, end = 1970)
pop70 <- popCalc(stLouisPop, start = 1970, end = 1980)

# combine with crime data
bind_rows(pop40, pop50, pop60, pop70) %>%
  distinct(year, population) %>%
  filter(year >= 1946) %>%
  left_join(stLouis, ., by = "year") %>%
  mutate(population.x = ifelse(is.na(population.y) == FALSE, population.y, population.x)) %>%
  select(-population.y) %>%
  rename(population = population.x) -> stLouis

# remove intermediate objects
rm(pop40, pop50, pop60, pop70, popYears, stLouisPop)
```

## Create Per Capita Rates
We are now able to calculate crime rates for murder, violent crime, property crime (only for a few years at this point), and thefts:

```{r}
stLouis %>%
  mutate(murderRate = (murder/population)*100000) %>%
  mutate(violentRate = (violentTotal/population)*100000) %>%
  mutate(propertyRate = (propertyTotal/population)*100000) %>%
  mutate(theftRate = (theftTotal/population)*100000) %>%
  select(year, population, violentTotal, violentRate, murder, murderRate,
         rape:propertyTotal, propertyRate, theftTotal, theftRate, 
         everything()) -> stLouis
```

## Write Data
Finally, data are written to `data/clean` for plotting and dissemination.

```{r}
write_csv(stLouis, here("data", "clean", "STL_CRIME_Trends.csv"))
```

